Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-02-28 analysis/paper/paper.tex
add S prefix to all tables and figs in supp mat

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
  pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\fancyhf{}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyheadoffset{0pt}
\rhead{\scshape A preprint - \today}
\cfoot{\thepage}

%Handling Keywords
\def\keywordname{{\bfseries \emph Keywords}}%
\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
}\noindent\keywordname\enspace\ignorespaces#1\par}}

% create title
% \providecommand{\maketitle}{}
% \renewcommand{\maketitle}{%
%   \par
%   \begingroup
%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
%     % for perfect author name centering
%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
%     % The footnote-mark was overlapping the footnote-text,
%     % added the following to fix this problem               (MK)
%     \long\def\@makefntext##1{%
%       \parindent 1em\noindent
%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
%     }
%     \thispagestyle{empty}
%     \@maketitle
%     \@thanks
%     %\@notice
%   \endgroup
%   \let\maketitle\relax
%   \let\thanks\relax
% }

% rules for title box at top of first page
% \newcommand{\@toptitlebar}{
%   \hrule height 2\p@
%   \vskip 0.25in
%   \vskip -\parskip%
% }
% \newcommand{\@bottomtitlebar}{
%   \vskip 0.29in
%   \vskip -\parskip
%   \hrule height 2\p@
%   \vskip 0.09in%
% }

% create title (includes both anonymized and non-anonymized versions)
% \providecommand{\@maketitle}{}
% \renewcommand{\@maketitle}{%
%   \vbox{%
%     \hsize\textwidth
%     \linewidth\hsize
%     \vskip 0.1in
%     \@toptitlebar
%     \centering
%     {\LARGE\sc \@title\par}
%     \@bottomtitlebar
%     \textsc{A Preprint}\\
%     \vskip 0.1in
%     \def\And{%
%       \end{tabular}\hfil\linebreak[0]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \def\AND{%
%       \end{tabular}\hfil\linebreak[4]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
%   }
% }

% add conference notice to bottom of first page
% \newcommand{\ftype@noticebox}{8}
% \newcommand{\@notice}{%
%   % give a bit of extra room back to authors on first page
%   \enlargethispage{2\baselineskip}%
%   \@float{noticebox}[b]%
%     \footnotesize\@noticestring%
%   \end@float%
% }

% abstract styling
\renewenvironment{abstract}
{
  \centerline
  {\large \bfseries \scshape Abstract}
  \begin{quote}
}
{
  \end{quote}
}

\usepackage{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy\(^1\), Amanda G. Henry\(^1\)}
\date{\(^1\) Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus has proven to contain a wealth of information
on the dietary habits of past populations. These insights have, to a large extent,
been obtained by the extraction and identification of
starch granules contained within the mineralised dental plaque from a wide range
of regions and time periods. The scope of previous studies have been limited to
microfossil extraction and identification to reconstruct dietary preferences from
the archaeological record, and few studies have attempted to address the biases
of starch retention in dental calculus. Those that have considered this problem
have been limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research on starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were exposed
to treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
and counted. We show that the granule counts extracted from the model dental calculus
represented a low proportion (ranging from 0.06\% to 0.16\%) of the total number
of granules exposed to the biofilms throughout the experiment.
Additionally, we found that the ratios of granule sizes from the extracted
starch granules differed from the original treatment solutions, with large
granules (\textgreater20 \(\mu\)m) consistently being under-represented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\)m in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}

@@

Dental calculus has proven to contain a wealth of
dietary information in the form of plant microfossils
(Hardy et al., 2009; Henry \& Piperno, 2008),
proteins (Hendy et al., 2018; Warinner, Hendy, et al., 2014),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within the mineralised dental plaque
over many millennia, providing a unique window into the food-related behaviours of
past populations
(Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Hardy et al., 2012; Henry et al., 2014).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

\begin{figure}

{\centering \includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see Figure \ref{fig:protocol-fig} for an overview of the protocol).

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

Statistical analysis was conducted in R version 4.1.1 (2021-08-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

@@

No \(\alpha\)-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
3.4 to 10.3 U/mL enzyme
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting \(\alpha\)-amylase activity
levels in humans, as the unit definition may differ;
however, they are sufficient to show that there is no activity in the system.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass[utf8]{frontiers/frontiersSCNS}

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% BELOW TAKEN FROM rticles plos template
%
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}
\usepackage{float}

\def\Authors{
  Bjørn Peare Bartholdy\,\textsuperscript{1*},
  Amanda G. Henry\,\textsuperscript{1}}

\def\Address{

  \textsuperscript{1} Archaeological Sciences, Leiden University,  Leiden,   Netherlands
  }

  \def\corrAuthor{Bjørn Peare Bartholdy}\def\corrAddress{Leiden University\\Einsteinweg 2\\Leiden, 2333 CC Netherlands}\def\corrEmail{\href{mailto:b.p.bartholdy@arch.leidenuniv.nl}{\nolinkurl{b.p.bartholdy@arch.leidenuniv.nl}}}
  \def\firstAuthorLast{BARTHOLDY {et~al.}}

\begin{document}

\onecolumn
\firstpage{1}

\title[Biases Associated with Starch Incorporation]{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

\tiny
 \keyFont{ \section{Keywords:} oral biofilm model, starch, alpha-amylase, dental calculus, paleodiet reconstruction}

\end{abstract}

@@

Dental calculus has proven to contain a wealth of
dietary information in the form of plant microfossils
(Henry and Piperno, 2008; Hardy et al., 2009),
proteins (Hendy et al., 2018; Warinner et al., 2014a),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within the mineralised dental plaque
over many millennia, providing a unique window into the food-related behaviours of
past populations
(Henry and Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Henry et al., 2014; Hardy et al., 2012).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner et al., 2014b),
microbotanical remains (Henry and Piperno, 2008; Hardy et al., 2009; Mickleburgh and Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Piperno and Dillehay, 2008; Henry et al., 2014; Tao et al., 2020; Jovanović et al., 2021).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce and Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see Figure \ref{fig:protocol-fig} for an overview of the protocol).

@@

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 \(\mu\)mole of maltose in 6 minutes.
The detailed protocol can be found here: \url{https://www.protocols.io/view/amylase-activity-bw8jphun}.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

Statistical analysis was conducted in R version 4.1.2 (2021-11-01) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

@@

No \(\alpha\)-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
(saliva)
contained amylase activity that could be detected in the assay, ranging from
9.93 to 30.2 U/mL enzyme
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting \(\alpha\)-amylase activity
levels in humans, as the unit definition may differ;
however, they are sufficient to show that there is no activity in the model.

@@

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.06\% to 0.16\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.06\% to 0.16\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp and Dudgeon, 2015; Tromp et al., 2017; Wesolowski et al., 2010), which could in part be due to the
lack of oral amylase activity in our model.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot}A);
a result that contradicts findings from archaeological contexts
(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.
